S4 Object-Oriented Programming
==============================
Here I'll keep notes about OOP using the S4 object system.

Classes and Instances/Objects
-----------------------------
A class is created using `setClass`. This takes the following key arguments:

* A __name__ for the class;
* A named list of __slots__ for the class;
* A class that it __contains__ i.e. inherits from.
* A default __prototype__ for the class;


```{r}
setClass("Person",
	 slots = list(
	   name = "character", age = "numeric"),
	 prototype = list(
	   name = NA_character_, age = NA_real_)
	 )

# Employee inherits Person
setClass("Employee",
	 slots = list(boss = "Person"),
	 contains = "Person")
getSlots("Person")
getSlots("Employee")
```

Create a new object of a class:

```{r}
nat <- new("Person", name = "Natalie", age = 36)
peter <- new("Employee", name = "Peter", age = 30, boss = nat)
nat
peter
```

Generic functions and methods
------------------------------
In order to define a method for a specific class, you need to make sure that a
generic method is defined for your method.
But how can you examine if a method has a generic?

* All primitive functions (e.g. check `getGenerics()`) have implicit generics;
* Use `isGeneric()`;
* Use `show()`;

```{r, error=TRUE}
getGenerics()
isGeneric("age")
show(age)
isGeneric("name")
show(name)
```

```{r}
setGeneric("age", function(obj, ...) standardGeneric("age"))
setMethod("age", "Person", function(obj, ...) obj@age)
setGeneric("name", function(obj, ...) standardGeneric("name"))
setMethod("name", "Person", function(obj, ...) obj@name)
isGeneric("age")
show(age)
isGeneric("name")
show(name)
age(nat)
name(nat)
str(nat)
```

Validity methods
----------------
You can define methods that apply additional restrictions for the validity of an object:

```{r}
check_person <- function(object) {
  errors <- character()
  length_age <- length(object@age)
  if (length_age != 1) {
    msg <- paste0("Age is length ", length_age, ". Should be 1.")
    errors <- c(errors, msg)
  }
  
  length_name <- length(object@name)
  if (length_name != 1) {
    msg <- paste0("Name is length ", length_name, ". Should be 1.")
    errors <- c(errors, msg)
    
  }
  
  if (length(errors) == 0) TRUE else errors
}
```

Let's try it out:

```{r, error=TRUE}
setClass("Person",
         slots = list(name = "character",
                      age = "numeric"),
         validity = check_person)
new("Person", name = "Hadley")
new("Person", name = "Hadley", age = 1:10)
```

But note that the check is not automatically applied when we modify slots directly:

```{r}
hadley <- new("Person", name = "Hadley", age = 31)
hadley@age <- 1:10
hadley
```

You can force a check with `validObject`:

```{r, error=TRUE}
validObject(hadley)

```

Accessor methods
----------------
You can access a slot of an object using `@` or `slot`:

```{r, error=TRUE}
peter@age
slot(peter, "age")
nat@age
slot(nat, "age")
age(nat) # not defined yet
```

But this is not an efficient way of doing so. You should instead define
an accessor method (_getter_). 

Display Source Code
-------------------
Let's say I'm using the `QDNAseq` package, and I want to see what the `plot`
function does with objects of class `QDNAseqReadCounts`. I need to use the
`getMethod` function or the selectMethod if the class inherits from a parent
class:

```{r}
require(QDNAseq)
methods("plot")
showMethods("plot")
head(selectMethod("plot", signature = c("QDNAseqReadCounts", "missing")), n = 20)
```
