A practical tutorial on S4 programming
======================================

## Experiment
Say we have conducted a microarray experiment using `n` probes across `m`
individuals.

* We can represent the results of this experiment using a matrix
  of size `n` x `m` called `marray`. This matrix stores the intensity for
  probe `1, 2, ..., n` for sample `1, 2, ..., m`.
* The sample annotation can be described in a data.frame `samp_annot`
  with `m` rows and any number of columns.
* The probe annotation can be described in a data.frame
  `prob_annot` with `n` rows and any number of columns.


```{r}
set.seed(42)
n <- 10
m <- 6
intensities <- rnorm(n * m, mean = 10, sd = 5)
marray <- matrix(intensities, ncol = m)
marray

# sample annotation
samp_annot <- data.frame(sampleID = 1:m,
			 condition = rep(c("WT", "MUT"), each = 3))
rownames(samp_annot) <- colnames(marray) <- LETTERS[1:m]
samp_annot

# probe annotation
prob_annot <- data.frame(geneID = 1:n,
			  pathway = sample(LETTERS, n, replace = TRUE))
rownames(prob_annot) <- rownames(marray) <- paste0("probe", 1:n)
prob_annot

marray
```

Now let's add these three pieces to a list:

```{r}
maexp <- list(marray = marray,
	      samp_annot = samp_annot,
	      prob_annot = prob_annot)
rm(marray, samp_annot, prob_annot, intensities)
str(maexp)
ls()
```

Let's access and manipulate some of the elements of the `maexp` list:

```{r}
maexp$samp_annot # the whole data.frame
summary(maexp$marray[, "A"]) # the A sample
(wt <- maexp$samp_annot[, "condition"] == "WT") # WT samples
maexp$marray["probe8", wt] # probe8 for WT samples
maexp[["marray"]]["probe3", !wt] # probe3 for MUT samples
```

```{r}
boxplot(maexp$marray,
	main = paste0("Boxplot representing the intensity\n",
		      "distributions of the 10 probes for the",
		      " 6 samples"),
	xlab = "Sample", ylab = "Probe Intensity")
```

Now let's extract the first 10 probes for the first 3 samples:

```{r}
x <- 1:5
y <- 1:3
marray2 <- maexp$marray[x, y]
prob_annot2 <- maexp$prob_annot[x, ]
samp_annot2 <- maexp$samp_annot[y, ]
maexp2 <- list(marray = marray2,
	       samp_annot = samp_annot2,
	       prob_annot = prob_annot2)
rm(marray2, samp_annot2, prob_annot2)
str(maexp2)
```

As you can see it's a bit cumbersome and error-prone. There's a much better way
to manipulate "objects" like microarrays.


## Class Definition
We can define a class with the `setClass` function. It requires a name
(`MArray`) and a content structure. The different elements of an S4 class are
called slots. When defining the slots, we provide their names and classes as a
named vector or list. You need to create objects with exactly these types of
slots.

```{r}
MArray <- setClass("MArray",
		   slots = c(marray = "matrix",
			     samp_annot = "data.frame",
			     prob_annot = "data.frame"))
```

The `setClass` function returns a constructor that can be used to create an
instance (object) of the MArray class.


```{r}
MArray() # an empty object
# MArray(marray = 1:2) # not allowed since should be matrix
MArray(marray = matrix(1:2)) # now allowed
ma <- MArray(marray = maexp$marray,
	     samp_annot = maexp$samp_annot,
	     prob_annot = maexp$prob_annot
	     )
class(ma)
ma
```

We can access individual slots from an object using `@` in a similar manner to
`$` with lists:


```{r}
ma@samp_annot
```

## Methods
A generic function, or generic for short, is a function that dispatches methods
to their appropriate class-specific implementation. A method `foo` will work in
a way `X` with an object of class `A` and in a different way `Y` with an object
of class `B`.

We define methods using `setMethod`. Before we do so, we need to check if such
a method already exists. If it does, then a generic also exists and we can
proceed normally with the new method.
If it does not, then we first need to create the generic and
then proceed with the new method.

### `show` method
The `show` method allows to define a convenient and more compact view of an
object when we type its name in the console.

```{r}
show
isGeneric("show")
hasMethod("show")
showMethods("show")
args("show")
```

As there is already a `show` method, we can proceed with the new method
definition using the `setMethod` function. We need a couple of things first:

* the class of the object we're implementing the method for:
    - MArray passed via the `signature` argument;
* the argument names that are defined in the generic:
    - use `args("show")` or just `show`;
    - this gives a single argument called `object`;
    - this name will need to be used when writing the `definition` of our
      method;

```{r}
ma
setMethod("show",
	  signature = "MArray",
	  definition = function(object) {
	      cat("An object of class ", class(object), "\n", sep = "")
	      cat(" ", nrow(object@marray), " features by ",
		  ncol(object@marray), " samples. \n", sep = "")
	      invisible(NULL)
	  })
ma
```

### Accessor method
Let's create a method `marray` to access the `marray` slot from the `MArray`
class.

* First we create a new generic function with `setGeneric`. The name of the
  generic is defined, together with the name of the argument(s) that will have
  to be re-used when defining class-specific methods.

```{r}
setGeneric("marray", function(object, ...) standardGeneric("marray"))
```

* Now we can create the new method using `setMethod`:

```{r}
setMethod("marray", "MArray",
	  function(object) object@marray)

marray(ma)
```

* Let's do the same for the `samp_annot` and `prob_annot` slots:

```{r}
setGeneric("samp_annot", function(object, ...) standardGeneric("samp_annot"))
setMethod("samp_annot", "MArray",
	  function(object) object@samp_annot)
samp_annot(ma)

setGeneric("prob_annot", function(object, ...) standardGeneric("prob_annot"))
setMethod("prob_annot", "MArray",
	  function(object) object@prob_annot)
prob_annot(ma)
```

### Subset method
We can create a subset method for an `MArray` object using the `[` operator.
Let's first look at a couple of usage examples:

```{r}
letters[1:5]
`[`(letters, 1:5) # just like a function!
```

We can directly implement this subsetting operator for the `MArray` class since
primitive functions are automatically promoted to generics when a method of the
same name is defined.

Using `?[` we can see that the function takes four arguments: the `x` object,
the `i` and `j` indices and the `drop` logical. We can ignore the final
argument by using the "missing" value:

```{r}
setMethod("[", "MArray", 
	  function(x, i, j, drop = "missing") {
	      .marray <- x@marray[i, j]
	      .prob_annot <- x@prob_annot[i, ]
	      .samp_annot <- x@samp_annot[j, ]
	      MArray(marray = .marray,
		     prob_annot = .prob_annot,
		     samp_annot = .samp_annot)
	  })

ma[1:3, c(2, 5)] # first 3 probes of samples 2 and 5
```

### Validity method
As discussed earlier, the number of rows in the `marray` slot needs to be equal
to the number of rows in the `prob_annot` slot, while the number of columns
needs to be equal to the number of rows in the `samp_annot` slot. The
respective names also need to be identical:

```{r}
setValidity("MArray", function(object) {
		msg <- NULL
		valid <- TRUE
		if (nrow(marray(object)) != nrow(prob_annot(object))) {
		    valid <- FALSE
		    msg <- c(msg, paste0("Number of data and probe annotation",
				    " rows must be identical."))
		}
		if (ncol(marray(object)) != nrow(samp_annot(object))) {
		    valid <- FALSE
		    msg <- c(msg, paste0("Number of data columns and sample ",
					 "annotation rows must be identical."))
		}
		if (!identical(rownames(marray(object)),
			       rownames(prob_annot(object)))) {
		    valid <- FALSE
		    msg <- c(msg, paste0("Data and probe annotation row names",
					 " must be identical"))
		}
		if (!identical(colnames(marray(object)),
			       rownames(samp_annot(object)))) {
		    valid <- FALSE
		    msg <- c(msg, paste0("Data column names and ",
					 "sample annotation row names",
					 " must be identical"))
		}
		if (valid) TRUE else msg
	  })

validObject(ma)

```
How about creating an invalid object?

```{r error=TRUE}
MArray(marray = matrix(1:12, ncol = 3),
       prob_annot = prob_annot(ma),
       samp_annot = samp_annot(ma))
```

### Replacement method
How can we update the content of slots? We can do so directly, but this
violates the encapsulation of our data and makes it possible to break the
validity of an object.

```{r error=TRUE}
ma@marray <- 1
(broken <- ma)
broken@marray <- matrix(1:9, ncol = 3)
broken
validObject(broken)
```

We can use a special type of replacement method using the code
`slot(object)<-syntax`. You need to concatenate the name of the method and the
assignment operator. A method to replace the slot that can be accessed with the
`marray` accessor will be called `marray<-`. It requires at least two
arguments: the `object` to be updated and the replacement `value`.
And since there is no `marray<-` generic, we first need to define one:

```{r}
setGeneric("marray<-",
	   function(object, value) standardGeneric("marray<-"))
```

In the definition of the replacement method we need to check that the
user-specified `value` does not break the validity of `object`:

```{r}
setMethod("marray<-", "MArray",
	  function(object, value) {
	      object@marray <- value
	      if (validObject(object))
		  return(object)
	  })
```

Now we can try it out:

```{r error=TRUE}
tmp <- matrix(rnorm(n * m, 10, 5), ncol = m)
marray(ma) <- tmp
colnames(tmp) <- LETTERS[1:m]
rownames(tmp) <- paste0("probe", 1:n)
head(marray(ma), n = 2) # before
marray(ma) <- tmp
head(marray(ma), n = 2) # after
```

* Let's do the same for the `samp_annot` and `prob_annot` slots:

```{r}
# First the generics
setGeneric("prob_annot<-",
	   function(object, value) standardGeneric("prob_annot<-"))
setGeneric("samp_annot<-",
	   function(object, value) standardGeneric("samp_annot<-"))

# Now the methods
setMethod("prob_annot<-", "MArray",
	  function(object, value) {
	      object@prob_annot <- value
	      if (validObject(object))
		  return(object)
	  })
setMethod("samp_annot<-", "MArray",
	  function(object, value) {
	      object@samp_annot <- value
	      if (validObject(object))
		  return(object)
	  })

```

Now we can try them out:

```{r}
samp_annot(ma)$sex <- rep(c("M", "F"), 3)
samp_annot(ma)
```

### Dissection methods
We can find out more about a class without reading its source code by using the
following functions:

```{r}
slotNames(ma) 
getClass("MArray") # shows slots and their classes
showMethods("marray")
showMethods(classes = "MArray")
getMethod("marray", "MArray") # show method code
getMethod("prob_annot", "MArray") 
```


### Sources:
* https://github.com/lgatto/S4-tutorial
