A practical tutorial on S4 programming
======================================
* Author: Laurent Gatto
* Sources:
    * https://github.com/lgatto/TeachingMaterial
    * https://github.com/lgatto/roo
    * https://github.com/lgatto/S4-tutorial

## Microarrays
Say we have conducted a microarray experiment using `n` probes across `m`
individuals.

* We can represent the results of this experiment using a matrix
  of size `n` x `m` called `marray`. This matrix stores the intensity for
  probe `1, 2, ..., n` for sample `1, 2, ..., m`.
* The sample annotation can be described in a data.frame `samp_annot`
  with `m` rows and any number of columns.
* The probe annotation can be described in a data.frame
  `prob_annot` with `n` rows and any number of columns.


```{r}
n <- 10
m <- 6
intensities <- rnorm(n * m, mean = 10, sd = 5)
marray <- matrix(intensities, ncol = m)
marray

# sample annotation
samp_annot <- data.frame(sampleID = 1:m,
			 condition = rep(c("WT", "MUT"), each = 3))
rownames(samp_annot) <- colnames(marray) <- LETTERS[1:m]
samp_annot

# probe annotation
prob_annot <- data.frame(geneID = 1:n,
			  pathway = sample(LETTERS, n, replace = TRUE))
rownames(prob_annot) <- rownames(marray) <- paste0("probe", 1:n)
prob_annot

marray
```

Now let's add these three pieces to a list:

```{r}
maexp <- list(marray = marray,
	      samp_annot = samp_annot,
	      prob_annot = prob_annot)
rm(marray, samp_annot, prob_annot, intensities)
str(maexp)
ls()
```

Let's access and manipulate some of the elements of the `maexp` list:

```{r}
maexp$samp_annot # the whole data.frame
summary(maexp$marray[, "A"]) # the A sample
(wt <- maexp$samp_annot[, "condition"] == "WT") # WT samples
maexp$marray["probe8", wt] # probe8 for WT samples
maexp[["marray"]]["probe3", !wt] # probe3 for MUT samples
```

```{r}
boxplot(maexp$marray,
	main = paste0("Boxplot representing the intensity\n",
		      "distributions of the 10 probes for the",
		      " 6 samples"),
	xlab = "Sample", ylab = "Probe Intensity")
```

Now let's extract the first 10 probes for the first 3 samples:

```{r}
x <- 1:5
y <- 1:3
marray2 <- maexp$marray[x, y]
prob_annot2 <- maexp$prob_annot[x, ]
samp_annot2 <- maexp$samp_annot[y, ]
maexp2 <- list(marray = marray2,
	       samp_annot = samp_annot2,
	       prob_annot = prob_annot2)
rm(marray2, samp_annot2, prob_annot2)
str(maexp2)
```

As you can see it's a bit cumbersome and error-prone. There's a much better way
to manipulate "objects" like microarrays.


## Define the class MArray
We can define a class with the `setClass` function. It requires a name
(`MArray`) and a content structure. The different elements of an S4 class are
called slots. When defining the slots, we provide their names and classes as a
named vector or list. You need to create objects with exactly these types of
slots.

```{r}
MArray <- setClass("MArray",
		   slots = c(marray = "matrix",
			     samp_annot = "data.frame",
			     prob_annot = "data.frame"))
```

The `setClass` function returns a constructor that can be used to create an
instance (object) of the MArray class.


```{r}
MArray() # an empty object
# MArray(marray = 1:2) # not allowed since should be matrix
MArray(marray = matrix(1:2)) # not allowed since should be matrix
```


