Creating a `sequences` S4 package
===============================
This simple package will handle sequence data. It will be able to load a FASTA
file and based on sequence type (DNA or RNA) do some operations, like finding
the sequence length or reverse sequence.

Functions we would like to have include:

* `readFasta()`: read in a single sequence from a FASTA file;
* `id()`, `seq()`: return the ID and the actual sequence of the sequence;
* `rev()`: return the reverse DNA/RNA sequence;
* `length()`: return the DNA/RNA sequence length;
* `comp()`: return complementary DNA/RNA sequence;
* `transcribe()`: return RNA sequence for DNA sequence.


Naive `readFasta()`
-------------------
Loads the data from a FASTA file and somehow represents it in R.

A FASTA file has a header line starting with `>` and then a nucleotide sequence
(DNA: ATCG; RNA: AUCG).

```
> example dna sequence
agcatacgacgactacgacactacgacatcagacactacagactactac
gactacagacatcagacactacatatttacatcatcagagattatatta
acatcagacatcgacacatcatcatcagcatcat
```

```{r}
# Create a class
setClass("GenericSeq",
	 slots = c(id = "character",
		   sequence = "character",
		   alphabet = "character"))

# Create an object of this class
genseq <- new("GenericSeq",
	      id = "sequence name",
	      sequence = "AGATACCCCGAAACGA",
	      alphabet = c("A", "C", "G", "T"))
genseq
genseq@id
slot(genseq, "id")
```

```{r}
readFasta <- function(infile) {
    lines <- readLines(infile)
    header <- grep("^>", lines)
    if (length(header) > 1) {
	warning("Reading first sequence only.")
	lines <- lines[header[1]:(header[2]-1)]
	header <- header[1]
    }
    id <- sub("^> *", "", lines[header], perl = TRUE)
    sequence <- toupper(paste(lines[(header+1):length(lines)], collapse = ""))
    alphabet <- unique(strsplit(sequence, "")[[1]])
    return.value <- new("GenericSeq",
			id = id, sequence = sequence, alphabet = alphabet)
    return.value
}
```

Now let's read the FASTA file:

```{r}
s <- readFasta("3-sequences_input.fasta")
s
```

Generics and methods
--------------------
* If a generic `id()` **does not exist already**, we need to create it to be
  able to use our method.

```{r}
setGeneric("id", function(obj, ...) standardGeneric("id"))
setMethod("id", "GenericSeq", function(obj, ...) obj@id)
id(genseq)
```

* If a generic `rev()` **exists already**, we can then create a method for our
  class.

```{r}
isGeneric("rev")
show(rev) # show source code
```

Since the source code contains `UseMethod("rev")`, it is an S3 generic. So we
need to define a S4 generic **with exactly the same parameters**.

```{r}
setGeneric("rev", function(x) standardGeneric("rev"))
setMethod("rev", "GenericSeq", function(x)
	  paste(rev(unlist(strsplit(x@sequence, ""))), collapse = ""))
```

All primitive functions (e.g. `c()`) have implicit generics.

```{r}
rev(genseq)
rev(1:5)
showMethods("rev")
```

Getters and Setters
-------------------
You should not use `@` in your code to access slots. Instead, you should
create getter and setter methods for all slots you want to expose.

```{r}
setGeneric("id<-",
	   function(object, value) standardGeneric("id<-"))
setReplaceMethod("id",
		 signature(object = "GenericSeq",
			   value = "character"),
		 function(object, value) {
		     object@id <- value
		     return(object)
		 })
id(genseq)
id(genseq) <- "new sequence name"
id(genseq)
genseq
```

Let's try the following introspection ("dissection") methods:

```{r}
showMethods("rev")
getClass("GenericSeq")
slotNames(genseq)
getMethod("rev", "GenericSeq")
findMethods("rev")
isGeneric("rev")
```

Let's implement some more getter/setter methods:

```{r}
setGeneric("seq", function(obj, ...) standardGeneric("seq"))
setMethod("seq", "GenericSeq", function(obj, ...) obj@sequence)
seq(genseq)
```

```{r}
setGeneric("seq<-",
	   function(object, value) standardGeneric("seq<-"))
setReplaceMethod("seq",
		 signature = c(object = "GenericSeq",
			       value = "character"),
		 function(object, value) {
		     object@sequence <- value
		     return(object) })
genseq2 <- genseq
seq(genseq2) <- "AAAAACCCCGGGTTTT"
seq(genseq2)
genseq2

```
